Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-06-21 2015-turklang-kazdep/paper.tex
updates
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60949 72bbbca6-d526-0410-a7d9-f06f51895060

\begin{figure}
\begin{tiny}
\begin{alltt}
"<Елде>"
        "ел" n loc
"<ағылшын>"
        "ағылшын" n nom
"<үстемдігіне>"
        "үстемдік" n px3sp dat
"<қарсы>"
        "қарсы" post
"<жаңа>"
        "жаңа" adj
"<толқулар>"
        "толқу" n pl nom
"<басталды>"
        "баста" v tv pass ifi p3 pl
"<.>"
        "." sent
\end{alltt}
\end{tiny}
\caption{Example output after preprocessing and manual disambiguation. Each token is annotated for lemma,
   part of speech and morphosyntactic information.}
\end{figure}

@@

The copula (both \emph{е} and \emph{бол}) is a challenging problem for dependency 
analysis of Kazakh. The universal dependency guidelines state that the copula should
be the dependent of the lexical predicate. However, in many cases the copula in Kazakh
is found in embedded clauses ...

@@

\begin{figure}[htbp]
	\centering
	\begin{dependency}[theme = simple, font = \small]
		\begin{deptext}[column sep=0.08cm]
			Бразилия \& өз \& жерінде \& чемпионатты \& екі \& рет \& өткізген \& бесінші \& ел \& болды. \\
			\gmk{np.nom} \& \gmk{det} \& \gmk{nom.px3sp.loc} \& \gmk{n.acc} \& \gmk{num} \& \gmk{n} \& \gmk{v.tv.gpr} \& \gmk{num.ord} \& \gmk{n.nom} \& \gmk{v.iv.ifi} \\
			\tgloss{Brasil} \& \tgloss{own} \& \tgloss{in its land} \& \tgloss{championship} \& \tgloss{two} \& \tgloss{times} \& \tgloss{having held} \& \tgloss{fifth} \& \tgloss{country} \& \tgloss{became} \\
		\end{deptext}
		\depedge{3}{2}{\udlabel{det}}
		\depedge{7}{3}{\udlabel{nmod}}
		\depedge{7}{4}{\udlabel{obj}}
		\depedge{6}{5}{\udlabel{nummod}}
		\depedge{7}{6}{\udlabel{nmod}}
		\depedge{9}{7}{\udlabel{acl}}
		\depedge{9}{8}{\udlabel{amod}}
		\depedge{9}{10}{\udlabel{cop}}
		\deproot[edge unit distance=1.3ex]{9}{\udlabel{root}}
	\end{dependency}
	\caption{Verbal adjectives: verbal adjectives are the head of everything in their own clause and are an \udtag{acl} dependency of the noun they modify.}\label{fig:gpr}
\end{figure}

>>>>

%\begin{figure}
%\begin{tiny}
%\begin{alltt}
%"<Елде>"
%        "ел" n loc
%"<ағылшын>"
%        "ағылшын" n nom
%"<үстемдігіне>"
%        "үстемдік" n px3sp dat
%"<қарсы>"
%        "қарсы" post
%"<жаңа>"
%        "жаңа" adj
%"<толқулар>"
%        "толқу" n pl nom
%"<басталды>"
%        "баста" v tv pass ifi p3 pl
%"<.>"
%        "." sent
%\end{alltt}
%\end{tiny}
%\caption{Example output after preprocessing and manual disambiguation. Each token is annotated for lemma,
%   part of speech and morphosyntactic information.}
%\end{figure}

@@

The copula (both \emph{е-} and \emph{бол-}) is a challenging problem for dependency 
analysis of Kazakh. The universal dependency guidelines state that the copula should
be the dependent of the lexical predicate. However, in many cases the copula in Kazakh
is found in embedded clauses ...

@@

\begin{figure}[htbp]
	\centering
	\begin{dependency}[theme = simple, font = \small]
		\begin{deptext}[column sep=0.08cm]
			Бразилия \& өз \& жерінде \& чемпионатты \& екі \& рет \& өткізген \& бесінші \& ел \& болды. \\
			\gmk{np.nom} \& \gmk{det} \& \gmk{nom.px3sp.loc} \& \gmk{n.acc} \& \gmk{num} \& \gmk{n} \& \gmk{v.tv.gpr} \& \gmk{num.ord} \& \gmk{n.nom} \& \gmk{v.iv.ifi} \\
			\tgloss{Brazil} \& \tgloss{own} \& \tgloss{in its land} \& \tgloss{championship} \& \tgloss{two} \& \tgloss{times} \& \tgloss{having held} \& \tgloss{fifth} \& \tgloss{country} \& \tgloss{became} \\
		\end{deptext}
		\depedge{3}{2}{\udlabel{det}}
		\depedge{7}{3}{\udlabel{nmod}}
		\depedge{7}{4}{\udlabel{obj}}
		\depedge{6}{5}{\udlabel{nummod}}
		\depedge{7}{6}{\udlabel{nmod}}
		\depedge{9}{7}{\udlabel{acl}}
		\depedge{9}{8}{\udlabel{amod}}
		\depedge{9}{10}{\udlabel{cop}}
		\deproot[edge unit distance=1.3ex]{9}{\udlabel{root}}
	\end{dependency}
	\caption{Verbal adjectives: verbal adjectives are the head of everything in their own clause and are an \udtag{acl} dependency of the noun they modify.}\label{fig:gpr}
\end{figure}
